---
title: "Climate Change"
author: "Aleph Jimenez"
date: "2020-11-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BACKGROUND

The planet's surface temperature is increasing as greenhouse gas emissions increase, and this global warming and carbon cycle disruption is wreaking havoc on natural systems. Living systems that depend on current temperature, weather, currents and carbon balance are jeopardized, and human society will be forced to contend with widespread economic, social, political and environmental damage as the temperature continues to rise. We examine the relationship between global temperature changes, greenhouse gases and human carbon emissions using time series of actual atmospheric and ice core measurements from the National Oceanic and Atmospheric Administration (NOAA) and Carbon Dioxide Information Analysis Center (CDIAC).

A major determinant of Earth's temperature is the greenhouse effect. Many
gases (e.g. C02, CH4, N20) trap heat and reflect it towards the surface, 
preventing heat from escaping the atmosphere. The greenhouse effect is vital in 
keeping Earth at a warm enough temperature to sustain liquid water and life; 
however, changes in greenhouse gas levels can alter the temperature balance of 
the planet.

Greenhouse gases have changed over the course of human history, 
but how has CO2 varied over a longer time scale?

One way to differentiate natural CO2 oscillations from today's manmade CO2 
spike is by examining the rate of change of CO2. The planet is affected not 
only by the absolute concentration of CO2 but also by its rate of change. When
the rate of change is slow, living and nonliving systems have time to adapt to
new temperature and gas levels, but when the rate of change is fast, abrupt 
differences can overwhelm natural systems. How does the pace of natural CO2 
change differ from the current rate of change?



```{r, libraries and data, echo=FALSE}
# Libraries and Data Import
library(tidyverse)
library(gridExtra)
library(ggthemes)
library(Rcpp)
library(Amelia)
library(GGally)
library(dslabs)
data(temp_carbon)
data(greenhouse_gases)
data(historic_co2)
options(digits = 3)

```

## EXPLORATORY

### 1. TEMP_CARBON

Let's check that time vector doesn't have gaps.

```{r, carbon time gaps}
years <- temp_carbon$year
y <- integer(length(temp_carbon$year))
plot(years, y, "l")

sequence <- seq(1,268)
plot(sequence, years, 'l')   # rearrange time vector


# Verifying there's no NA's
sum(is.na(temp_carbon))        # [1] 391, missing data.
temp_carbon %>%  missmap(col = c("black", "grey"))

```



Checking variable correlations

```{r, carbon correlations}
ggcorr(temp_carbon,
       nbreaks = 6,
       label = TRUE,
       label_size = 3,
       color = "grey50")

```


Graphs of Global, Land and Ocean Temperatures Anomalies

```{r, graphs carbon anomalies}
Global <- temp_carbon$temp_anomaly
Land <- temp_carbon$land_anomaly
Ocean <- temp_carbon$ocean_anomaly

df <- data.frame(years, 
                 Global, 
                 Land, 
                 Ocean)

dfplot <- df %>% gather(key, value, -years)

ggplot(dfplot, mapping = aes(x = years, 
                             y = value, 
                             color = key)) + 
  geom_line(na.rm = TRUE) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 1938, linetype = "twodash") +
  scale_x_continuous(limits = c(1870, 2020),
                     breaks = c(1875, 1900, 1950, 2000, 2020)) +
  scale_color_discrete(name = "Anomalies") +
  xlab("Years") +
  ylab("Temperature [C]") +
  ggtitle("Annual Global, Land and Ocean Temperatures Anomalies") +
  theme(plot.title = element_text(hjust = 0.5))

```


Graph for Annual Carbon Emissions

```{r, graphs annual carbon}
yline <- temp_carbon$carbon_emissions[temp_carbon$year == 1938]

temp_carbon %>% 
  ggplot(aes(year, carbon_emissions)) + 
  geom_line(na.rm = TRUE, 
            color = "blue", 
            linetype = "twodash") +
  geom_hline(yintercept = 0) + 
  geom_hline(yintercept = yline, 
             color = "green", 
             linetype = "twodash") +
  geom_vline(xintercept = 1850, linetype = "solid") +
  geom_vline(xintercept = 1938, linetype = "twodash") +
  scale_y_log10(breaks = c(0, 10, 100, 1000, yline, 2500, 5000, 10000)) +
  scale_x_continuous(breaks = c(1800, 1850, 1900, 1938, 2000)) +
  xlab("Year") + 
  ylab("C6 Emission [metric tons/year]") +
  ggtitle("Annual Carbon Emissions") + 
  theme_bw()

```




### 2. GREENHOUSE_GASES

Let's check that time vector doesn't have gaps.
The time vector (year) has three time lines for each gas. First we run a code for
all vectors. If something is wrong we would run it for each gas

```{r, ghg time gaps}
years <- greenhouse_gases$year    
y <- integer(length(greenhouse_gases$year))   # making a vector of 0's; 
plot(years, y, "l")                           # normalizing with 100 for each gas at the end


# Verifying that there's no NA's
sum(is.na(greenhouse_gases))        # [1] 0, missing data.
greenhouse_gases %>%  missmap(col = c("black", "grey"))

```


Checking variables correlation

```{r, ghg correlations}
df <- greenhouse_gases %>%
  spread(gas, concentration)
  
ggcorr(df,
       nbreaks = 4,
       label = TRUE,
       label_size = 3,
       color = "grey50")
```


Graphs of Concentration of Greenhouse Gases

```{r, graphs ghg}
# All
# 1.1
greenhouse_gases %>% 
  mutate(gas = factor(gas)) %>%
  group_by(gas) %>% 
  ggplot(aes(year, concentration, color = gas)) + 
  geom_line() +
  scale_y_continuous(trans = "log10",
                     breaks = c(0, 200, 300, 400, 500, 600, 1000, 1500, 1800)) +
  scale_color_discrete(name = "Gases") +
  xlab("Year") +
  ylab("Concentration 
       (CO2[ppm]; CH4,N2O[ppb])") +
  ggtitle("Global Gas Concentration over 2000 years.")
  

# Each
# 1.2.1 CO2 & N2O
average <- greenhouse_gases %>%
  filter(year %in% seq(0, 1840, 20)) %>%      # 1850 unofficial year it begins 
                                              # the industrial revolution
  summarize(CH4 = mean(concentration[gas == "CH4"]),
            CO2 = mean(concentration[gas == "CO2"]), 
            N2O = mean(concentration[gas == "N2O"]))

greenhouse_gases %>% 
  filter(gas %in% c("CO2", "N2O")) %>%
  mutate(gas = factor(gas)) %>%
  group_by(gas) %>%
  ggplot(aes(year, concentration, color = gas)) + 
  geom_line() +
  geom_hline(yintercept = average$CO2, linetype = "dashed") +
  geom_hline(yintercept = average$N2O, linetype = "dotted") +
  scale_y_continuous(limits = c(250, 375), 
                     breaks = c(250, 267, 279, 300, 325, 350, 375)) +
  scale_color_discrete(name = "Gases") +
  xlab("Year") +
  ylab("Concentration 
       CO2[ppm], N2O[ppb]") +
  ggtitle("Global Gas Concentration of CO2 and N2O over 2000 years.") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(x = 500, y = 282, 
           label = "CO2 average", colour = "blue", size = 3,
           show.legend = FALSE) +
  geom_text(x = 500, y = 270, 
            label = "N2O average", colour = "blue", size = 3,
            show.legend = FALSE)



# 1.2.2 CH4
greenhouse_gases %>% 
  filter(gas == c("CH4")) %>%
  ggplot(aes(year, concentration)) + 
  geom_line(color = "red") +
  geom_hline(yintercept = average$CH4, linetype = "dashed") +
  scale_y_continuous(limits = c(600, 1710), 
                     breaks = c(600, 662, 750, 1000, 1250, 1500, 1750)) +
  xlab("Year") +
  ylab("Concentration CH4[ppb]") +
  ggtitle("Global Gas Concentration of CH4 over 2000 years.") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(x = 500, y = 690, 
            label = "CH4 average", colour = "blue", size = 3.5,
            show.legend = FALSE)

```



### 3. HISTORIC_CO2

Checking that time vector doesn't have gaps.

```{r, historic co2 time gaps}
years <- historic_co2$year/1000            # time is in thousands years.
y <- integer(length(historic_co2$year))
plot(years, y, "l")

sequence <- seq(1, 694)
plot(sequence, years, 'l')   


# Verifying that there's no NA's
sum(is.na(historic_co2))        # [1] 0, missing data.
historic_co2 %>%  missmap(col = c("black", "grey"))

```


Checking variables correlation

```{r, historic co2 correlations}
df <- historic_co2 %>% 
  spread(source, co2)

ggcorr(df,
       nbreaks = 3,
       label = TRUE,
       label_size = 3,
       color = "grey50")

```



Graphs of Concentration of Historic CO2.

```{r, graphs historic co2}
# 1.1 Ice Cores and Mauna Loa data in one graph one figure.
historic_co2 %>%
  mutate(source = factor(source)) %>%
  group_by(source) %>%
  ggplot(aes(year/1000, co2, color = source)) +
  geom_line() +
  scale_color_discrete(name = "Source") +
  xlab("Year (10^3)") +
  ylab("CO2[ppm]") +
  ggtitle("Historic CO2 record from > 800,000 years")
  

# 1.2 Ice Cores and Mauna Loa data in two graphs one figure ('facet_grid').
historic_co2 %>%
  mutate(source = factor(source)) %>%
  group_by(source) %>%
  ggplot(aes(year, co2)) +
  geom_line() +
  facet_grid(source ~ ., scales = "free")


# 1.3 Ice Cores and Mauna Loa data in two graphs one figure ('grid.arrange').
avg_ic <- historic_co2 %>%
  filter(year < 1850 & source == "Ice Cores") %>%
  summarize(CO2 = mean(co2))

avg_ml <- historic_co2 %>%
  filter(year >= 1850 & source == "Mauna Loa") %>%
  summarize(CO2 = mean(co2))

p1 <- historic_co2 %>%
  filter(source == "Ice Cores") %>%
  ggplot(aes(year/1000, co2)) +
  geom_line(color = "blue") +
  geom_hline(yintercept = avg_ic$CO2) +
  geom_hline(yintercept = 368, linetype = "longdash") +     # 368 ppm, max of CO2 Ice Cores
  geom_vline(xintercept = 1850/10^3, linetype = "dashed") +
  geom_vline(xintercept = 1959/10^3, linetype = "longdash") +
  scale_y_continuous(breaks = c(200, avg_ic$CO2, 250, 300, 350, 368)) +
  xlab("Year*(10^3)") +
  ylab("Ice Cores CO2[ppm]") +
  ggtitle("Historic CO2 record > 800,000 from Ice Cores")

p2 <- historic_co2 %>%
  filter(source == "Mauna Loa") %>%
  ggplot(aes(year, co2)) +
  geom_line(color = "red") +
  geom_point() +
  geom_hline(yintercept = avg_ml$CO2) +
  geom_hline(yintercept = 368, linetype = "longdash") +
  xlab("Year") +
  ylab("Mauna Loa CO2[ppm]") +
  ggtitle("Historic CO2 record from Mauna Loa") +
  geom_text(x = 1970, y = 360, 
            label = "CO2[ppm] = 354", 
            color = "blue",
            size = 3.5)

grid.arrange(p1, p2, nrow = 2)


# 1.4.1 Ice Cores and Grenhouse CO2 data base comparison (one graph).
df_ic <- data.frame(year = historic_co2$year[historic_co2$year > 0 & 
                                             historic_co2$source == "Ice Cores"],
                    CO2 = historic_co2$co2[historic_co2$year > 0 & 
                                             historic_co2$source == "Ice Cores"])

df_gh <- data.frame(year = greenhouse_gases$year[greenhouse_gases$gas == "CO2"], 
                    CO2 = greenhouse_gases$concentration[greenhouse_gases$gas == "CO2"])


df <- rbind(df_ic, df_gh)

df$dataset <- c(rep("IC", nrow(df_ic)), rep("GH", nrow(df_gh)))

avg <- (mean(df_ic$CO2[df_ic$year < 1850]) + 
          mean(df_gh$CO2[df_gh$year < 1850]))/2

ggplot(data = df, aes(year, CO2, col = dataset)) + 
  geom_line() + 
  geom_point() +
  geom_hline(yintercept = avg) +
  geom_vline(xintercept = 1850, linetype = "dashed") +
  geom_vline(xintercept = 1938, linetype = "dashed") +
  scale_x_continuous(breaks = c(0, 500, 1000, 1500, 1850, 2000)) +
  xlab("Year") +
  ylab("CO2[ppm]") +
  scale_color_discrete(name = "Source") +
  ggtitle("Comparison between CO2 from Greenhouse gases(GH) 
          and Ice Cores(IC)")


# 1.4.2 (two graph)
p <- historic_co2 %>%        
  filter(year > 0 & source == "Ice Cores") %>%
  ggplot(aes(year, co2)) +            # length 85
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 1850, linetype = "dashed") +
  geom_vline(xintercept = 1938, linetype = "dashed") +
  ylab("Ice Cores CO2")

q <- greenhouse_gases %>%     
  filter(gas == "CO2") %>% 
  ggplot(aes(year, concentration)) +  # length 100
  geom_line() + 
  geom_point() +
  geom_vline(xintercept = 1850, linetype = "dashed") +
  geom_vline(xintercept = 1938, linetype = "dashed") +
  ylab("Greenhouse CO2")

grid.arrange(p, q, nrow = 2)


# 1.5 Ice Cores and Mauna Loa data from 1959.
historic_co2 %>%
  filter(year >= 1959) %>%
  ggplot(aes(year, co2, color = source)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 1959, linetype = "dashed") +
  xlab("Year") +
  ylab("CO2[ppm]") +
  ggtitle("CO2 from Ice Cores and Mauna Loa")
  
```


